---
title: "The exactmed functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The exactmed functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  prompt = TRUE,
  comment = " "
)
```

## Introduction

This document aims to illustrate the usage of the functions `exactmed()` and `exactmed_c()`, as well as their behavior via additional examples.

 We recall that `exactmed()` computes natural direct and indirect effects, and controlled direct effects for a binary outcome and a binary mediator. The user can specify the reference levels of the outcome and mediator variables using the input parameters `hvalue_m` and `hvalue_y`, respectively (see the function help). Controlled direct effects are obtained for both possible mediator values ($m=0$ and $m=1$). Natural and controlled effects can either be unadjusted or adjusted for covariates (that is, conditional effects). Adjusted effect estimates are obtained for covariates fixed at their sample-specific mean values (default) or at specific values (user-provided). Also, by default, `exactmed()` incorporates a mediator-exposure interaction term in the outcome model, which  can be removed by setting `interaction=FALSE`. Concerning interval estimates, `exactmed()`  generates, by default, $95\%$ confidence intervals obtained by the delta method. Alternatively, bootstrap confidence intervals, instead of delta method confidence intervals, can be obtained by specifying `boot=TRUE` in the function call. In this case, 1000 bootstrap datasets are generated by default.
 
The function `exactmed_c()` computes natural direct and indirect effects, and controlled direct effects for a binary outcome and a *continuous* mediator. 

This function shares almost all of the input parameters present in the first one. However, there are two small differences worth mentioning. 
One is the absence of the `hvalue_m` parameter of the `exactmed()` function because the mediator is now a continuous variable. The other is the addition of a new input parameter called `mf`; this parameter allows the user to specify the value at which the mediation variable is set in the calculation of the controlled direct effect. 

Due to the similarity of the two functions in terms of use and options offered to the user, the different examples will be presented with the `exactmed()` function. Thus, the reader interested in the `exactmed_c()` function for the continuous case will only need to change the name of the function (and the dataset) in the calling examples to understand its use.  That said, at the end of the document we present the dataset of the package called `datamed_c`  available for the continuous case and a couple of calling examples.
 
 In all the `exactmed()` examples presented below we use the dataset `datamed`, available after loading the **ExactMed** package. Some of the features of this dataset can be found in its corresponding help file (`help(datamed)`). We recall that the `exactmed()` function only works on data frames with named columns and no missing values.
 
```{r}
library(ExactMed)

head(datamed)

```
 
The following command verifies whether the dataset contains any missing values:
```{r}

as.logical(sum(is.na(datamed)))

```
 


## Basic examples
 
Suppose that one wishes to obtain mediation effects estimates for a change in exposure from $0$ to $1$, assuming there is no exposure-mediator interaction and using the delta method to construct $95\%$ confidence intervals.

In this case, a valid call to `exactmed()` would be:
 

```{r}

results1 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', 
  a1 = 1, a0 = 0, interaction = FALSE
  )  

results1

```
 
 
Mediation effects estimates adjusted for covariates are obtained through the use of the character vectors `m_cov` and `y_cov`, which contain the names of the covariates to be adjusted for in the mediator and outcome models, respectively. The following call to `exactmed()` incorporates covariates `C1` and `C2` in both the mediator and outcome models:

```{r}

results2 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0,  
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  interaction = FALSE
  )

results2

```


`exactmed()` also allows for the specification of two different sets of covariates in the mediator and outcome models. For example, the following specification of `m_cov` and `y_cov` means that the mediator model is adjusted for `C1` and `C2`, while the outcome model is adjusted for `C1` only.

However, we advise against this practice unless it is known that excluded covariates are independent of the dependent variable (mediator or outcome) being modeled given the rest of covariates.  


```{r}

results3 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0,  
  m_cov = c('C1', 'C2'), y_cov = c('C1'), 
  interaction = FALSE
  )

results3

```


By default, the `adjusted` parameter is `TRUE`. If the `adjusted` parameter is set to `FALSE`, `exactmed()` ignores the values of the vectors `m_cov` and `y_cov`, and computes unadjusted (crude) effect estimates as in the first example above: 
 
```{r}

results4 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1'), 
  adjusted = FALSE, interaction = FALSE
  )

results4

```
 

To perform an adjusted mediation analysis allowing for exposure-mediator interaction (by default `TRUE`) and using bootstrap based on $100$ resamples with initial random seed $= 1991$ to construct $97\%$ confidence intervals, one should call `exactmed()` as follows:

 
```{r, results='hide'}

results5 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  boot = TRUE, nboot = 100, bootseed = 1991, confcoef = 0.97
  )

```
 
 
```{r}

results5

```


## Firth's penalization 

 
In the situation where we believe that we are facing a problem of separation or quasi-separation, Firth's penalization can be used by setting the `Firth` parameter to `TRUE` (Firth penalized mediation analysis).
If this is the case, Firth's penalization is applied to both the mediator model and the outcome model.  


The `Firth` parameter implements Firth's penalization to reduce the bias of the regression coefficients estimators under scarce or sparse data:

```{r, results='hide'}

results6 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), Firth = TRUE, 
  boot = TRUE, nboot = 100, bootseed = 1991, confcoef = 0.97
  )

```

 
```{r}

results6

```
 
 
## Stratum-specific effects

 
The following call to `exactmed()` returns mediation effects when the values of the covariates `C1` and `C2` are $0.1$ and $0.4$, respectively, assuming an exposure-mediator interaction and using the delta method to construct $95\%$ confidence intervals:
 
 
```{r}

results7 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1,C2 = 0.4), y_cov_cond = c(C1 = 0.1, C2 = 0.4)
  )

results7

```
 
 
 
Common adjustment covariates in vectors `m_cov` and `y_cov` must have the same values; otherwise, the execution of the `exactmed()` function is aborted and an error message is displayed in the R console. Example:


```{r, error=TRUE, collapse=FALSE}

exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1','C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.3, C2 = 0.4), y_cov_cond = c(C1 = 0.1, C2 = 0.4)
 )


```


If the covariates specified in `m_cov_cond` (`y_cov_cond`) constitute some proper subset of `m_cov` (`y_cov`) then the other covariates are set to their sample-specific mean levels. Hence, the call

```{r}

results8 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1), y_cov_cond = c(C1 = 0.1)
  )

```

 
 is equivalent to:
 
```{r}

 mc2 <- mean(datamed$C2)
 mc2

results9 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1, C2 = mc2), y_cov_cond = c(C1 = 0.1, C2 = mc2)
  )

```
 
 
This can be checked by comparing the two outputs:
 
 
```{r}

identical(results8, results9)

```
 

With this in mind, an error is easily predicted if one makes this call:


```{r,error=TRUE, collapse=FALSE}

exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1), y_cov_cond = c(C1 = 0.1, C2 = 0.4)
  )


```

 
 
## Categorical covariates

`exactmed()` also allows for categorical covariates. Covariates of this type must appear in the data frame as factor, character or logical columns. To illustrate how `exactmed()` works with categorical covariates, we replace the covariate `C1` in the dataset `datamed` by a  random factor column:

```{r}

cate <- factor(sample(c("a", "b", "c"), nrow(datamed), replace =TRUE))
datamed$C1 <- cate

```


It is possible to estimate mediation effects at specific values of categorical covariates using the input parameters `m_cov_cond` and `y_cov_cond`. Note that if the targeted covariates are both numerical and categorical, the above parameters require to be list-type vectors, instead of atomic vectors as when covariates are only numerical.
 
 Hence, if one wants to estimate mediation effects at level 'a' for `C1` and at value $0.4$ for `C2`, assuming an exposure-mediator interaction and using the delta method to construct $95\%$ confidence intervals, `exactmed()` should be called as follows:
 
 
```{r}

results10 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = list(C1 = 'a', C2 = 0.4), y_cov_cond = list(C1 = 'a', C2 = 0.4)
  )

results10

```
 

If one does not specify a value for the categorical covariate `C1`, `exactmed()` computes the effects by assigning each dummy variable (created internally by `exactmed()`) to a value equal to the proportion of observations in the corresponding category (equivalent to setting each dummy variable to its mean value): 


```{r}

results11 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C2 = 0.4), y_cov_cond = c(C2 = 0.4)
  )

results11

```
 
 
## Case Control Mediation 

`exactmed()`  can also compute mediation effects in the context of a case-control study where response and mediator are binary. To do so, the true outcome prevalence has to be known and the user has to set the 
`yprevalence` parameter to this value. 
`exactmed()` accounts for the ascertainment in the sample employing weighted regression techniques that uses inverse-probability weighting (IPW) with robust standard errors.

The following call to `exactmed()` returns mediation effects supposing that the data has been obtained from a case-control study  and that the outcome prevalence is $0.1$:

```{r}

results12 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', 
  a1 = 1, a0 = 0, interaction = FALSE, yprevalence = 0.1
  )

results12

```

It is worth mentioning that, in the context of case control mediation, the same optional parameters are available to the user as described in the previous sections.


##  Mediation with a continous mediator

As mentioned in the introduction, in the case of a continuous mediator, the **ExactMed** package allows the user to obtain estimates of the different mediation effects by means of the `exactmed_c()` function, which offers practically the same options as `exactmed()`.
 The only difference is the absence of the `hvalue_m` parameter and the incorporation of the `mf` parameter, that allows setting a value of the mediator for the calculation of the controlled direct effect (by default `mf=1`).

The package also makes available to the user the `datamed_c` dataset containing a continuous mediation variable. Some of the features of this dataset can be found in its corresponding help file (`help(datamed_c)`). We recall that the `exactmed_c()` function only works on data frames with named columns and no missing values.

```{r}

library(ExactMed)

head(datamed_c)


```

Example of call to `exactmed_c()` in order to obtain estimates of conditional effects supposing no exposure-mediator interaction in the outcome regression model. 

```{r}

results13 <- exactmed_c(
  data = datamed_c, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0,  
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  interaction = FALSE
  )

results13

```


To perform an adjusted mediation analysis allowing for exposure-mediator interaction, using bootstrap based on $100$ resamples with initial random seed $= 1885$ to construct $95\%$ confidence intervals and computing controlled direct effect at the value $2$ of the mediator, one should call `exactmed_c()` as follows:

 
```{r, results='hide'}

results14 <- exactmed_c(
  data = datamed_c, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  boot = TRUE, nboot = 100, bootseed = 1885, confcoef = 0.95,
  mf = 2
  )

```
 
 
```{r}

results14

```

 
 
 
 

